---
title: "R Basics – SPC 2025 Vanuatu Training"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
author: "Your Name"
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This document introduces some fundamental R concepts that will help you work with data effectively. We'll cover libraries, assignment, functions, dataframes, pipes, filtering, and selecting data.

You can run the code chunks in this document using RStudio. Just click the green play button at the top right of each chunk.

You can also run individual lines of code by placing your cursor on the line (or by selecting multiple lines) and pressing `Ctrl + Enter` (or `Cmd + Enter` on Mac).

## 1. Libraries

You can think of libraries as toolboxes that contain pre-built functions to help you do specific tasks.

In [this](https://support.posit.co/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages) post you can find a list of useful R packages.

```{r}
# Packages of tools others have built
library(dplyr)   # Data manipulation
library(stringr) # Text handling
library(DT)      # Interactive tables
```

## 2. Assignment `<-`

```{r}
# Saving results to use later
my_numbers <- c(5, 10, 15, 20, 25)
## check the content of my_numbers

# Q: what if I want to create a vector of characters instead?
```

## 3. Functions

```{r}
mean(my_numbers)
max(my_numbers)

# Q: what if I want to calculate the minimum value?
```

## 4. Dataframe and Pipes `|>`

Chains commands together (like "then do this")

```{r, echo = FALSE}
# Sample data frame
data <- data.frame(
  vessel = c("A", "B", "C", "D"),
  year = c(2023, 2024, 2024, 2023),
  catch = c(100, 150, 200, 250)
)

data |>
  head(2) |>
  DT::datatable()

# Q: what if I want to display the first 3 rows instead?
```

### 4.1 Read csv to a dataframe and run summarisation (group by)
```{r, echo = FALSE}
# load csv data to R (as a dataframe)
csv_data <- read.csv("data/vu_3397.csv")

# what if I want to see the column names?
csv_data |>
  colnames()
## open the df from the global environment to see its structure

# what if I want to see distinct yearmonths?
csv_data |>
  distinct(Yearmonth)

# what if I want to see distinct vessels and yearmonths?
csv_data |>
  distinct(Vessels, Yearmonth)

# what if I want to see the total n of trips by vessel?
csv_data |>
  group_by(Vessels) |>
  summarise(total_n_trips = sum(Trips, na.rm = TRUE)) |>
  arrange(desc(total_n_trips)) |>
  DT::datatable()

# Q: and what about if I want to know the total number of trips by vessel and yearmonth?


```



## 5. Filter and select

Chains commands together (like "then do this") and use filtering commands

```{r}
data |>
  filter(year == 2024) |> # select only 2024 data
  select(vessel, catch)   # select specific columns

# Q: what if I want to filter by year 2023 and select vessel and year columns?
```

**Reads like:** "Take data, THEN filter by year, THEN select columns"
